The Precambrian paleogeography of Laurentia

### 2020-09-22 Manuscript/Laurentia_Paleogeo_Manuscript.tex
add "Associations prior to Laurentia assembly" subsection

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989b, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989b}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989b} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}. TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}. Bottom panel: velocity implied by the paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
\begin{itemize}
%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}. The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited 
\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preserved within secondary silica \citep{White1984a}. The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}. Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}. Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}. There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.  These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 Ma within the Helena Formation \citep{Evans2000c}. 
\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.  These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} . These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca. 1005 Ma detrital zircons \citep{Turner2016a}. The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca. 820 Ma \citep{Turner2016a}. These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in an intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}. The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}. The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation \citep{Rooney2013a}.
\end{itemize}

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}
Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia. This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models.
\textit{Note that this section is currently under development and is incomplete}

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers. This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013b, Cawood2017a}. While the lack of ca. 1100 and ca. 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; D'Agrella-Filho et al., this volume) as shown in \cite{Evans2013a}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.

>>>>

\subsection{Introduction and broad tectonic history}
\lettrine[lines=2]{L}{aurentia} refers to the craton that forms the Precambrian interior of North America and Greenland (Fig. \ref{fig:Laurentia_map}). Laurentia comprises multiple Archean provinces that had unique histories prior to their amalgamation in the Paleoproterozoic (ca. 1.8 Ga), as well as tectonic zones of crustal growth that post-date this assembly \citep{Hoffman1989c, Whitmeyer2007a}. Collision between the Superior province and the composite Slave+Rae+Hearne provinces that resulted in the Trans-Hudson orogeny represents a major event in the formation of Laurentia \citep{Corrigan2009a}. Terminal collision recorded in the Trans-Hudson orogen is estimated to have been ca. 1.86 to 1.82 Ga based on constraints such as U-Pb dating of monazite grains and zircon rims \citep{Skipton2016a, Weller2017a}. A period of accretionary and collisional orogenesis is recorded in the constituent provinces and terranes of Laurentia leading up to the terminal collision of the Trans-Hudson orogeny. This overall story of rapid Paleoproterozoic amalgamation of Laurentia's constituent Archean provinces, including the terminal Trans-Hudson orogeny, was synthesized in the seminal \textit{United Plates of America} paper of \citet{Hoffman1988a} and has been refined in the time since -- particularly with additional geochronological constraints. Of most relevance here are the events that led to the suturing of more major Archean provinces: the Thelon orogen associated with the collision between the Slave and Rae provinces ca. 2.0 to 1.9 Ga \citep{Hoffman1989c}; the Snowbird orogen associated with ca. 1.90 Ga collision between the Rae and Hearne provinces and associated terranes \citep{Berman2007a, Thiessen2020a}; the Nagssugtoqidian orogen due to the ca. 1.86 to 1.84 Ga collision between the Rae and North Atlantic provinces \citep{St-Onge2009a}; and the Torngat orogen resulting from the ca. 1.87 to 1.85 Ga collision of the southern Meta Incognita province (grouped with the Rae province in older compilations) with the North Atlantic province \citep{St-Onge2009a}.

@@

There is significantly less preserved Mesoproterozoic crustal growth on the western margin of Laurentia (Fig. \ref{fig:Laurentia_map}), and the  tectonic history through the Mesoproterozoic Era is not as well elucidated as on the southern to eastern margin. The 15 to 20 km thick package of sedimentary rocks of the Belt-Purcell Supergroup is associated with a ca. 1.47 to 1.40 rift \citep{Evans2000c}. While the rift is typically interpreted as being intracontinental \citep{Lydon2004a}, the tectonic setting in which it formed is debated. \citet{Hoffman1989c} proposed that it may be a remnant back-arc basin trapped within a continent, while others envision it as being associated with continental rifting along the margin associated with separation of a conjugate continent \citep{Jones2015a}. This region is interpreted to have been subsequently deformed during a ca. 1.36 to 1.33 Ga event known as the East Kootenay orogeny \citep{McMechan1982a, Nesheim2012a, McFarlane2015a}.

@@

\begin{figure*}
\centering
\includegraphics[width=0.9\textwidth]{../Figures/Laurentia_paleolatitude.pdf}
\caption{\textbf{Laurentia paleolatitude through time in data and models.} Top panel: Paleolatitude implied by paleomagnetic poles from Laurentia and associated blocks for Duluth (lat=46.79\textdegree N, lon=92.10\textdegree W). The paleomagnetic poles are compiled in Table 2. Middle panel: Paleolatitude implied by Laurentia poles compared with that implied by published paleogeographic models and the simple Laurentia model used in this chapter for the reconstructions in Figure \ref{fig:Laurentia_reconstructions}. TC2017 refers to \cite{Torsvik2017a} and S-H 2017 refers to \cite{Swanson-Hysell2017a}. Bottom panel: velocity implied by the continuous paleogeographic models in cm per year for the Duluth reference point on Laurentia.}
\label{fig:Laurentia_paleolatitude}
\end{figure*}

@@

Proterozoic evaporite deposits are documented within these units that were deposited following the amalgamation of Laurentia: 
\begin{itemize}
%\item The Stark Formation of the Slave Province contains displacive halite pseudomorphs throughout as well as massive breccias  interpreted to be evaporite-collapse breccias \citep{Pope2003a}. The formation is intruded by 1865 $\pm$ 15 Compton laccoliths  and is interpreted to have been deposited 
\item The Altyn Formation of the Belt Supergroup contains pseudomorphs after gypsum crystals and anhydrite within shallow-water carbonates with relict gypsum and anhydrite preserved within secondary silica \citep{White1984a}. The correlative Prichard Formation is intruded by 1468.8 $\pm$ 2.5 Ma sills \citep{Sears1998a}. Halite molds and casts are present within mudstones of the overlying Grinell Formation \citep{Pratt2019a}. Higher in the Belt Supergroup stratigraphy, within the Wallace Formation, there is stratiform scapolite --- a metamorphic mineral interpreted to have formed from a halite precursor within the Wallace Formation \citep{Hietanen1967a}. There are also halite and gympsum pseudomorphs within carbonate mudstones of the correlative to underlying Helena Formation \citep{Pratt2001a,Winston2007a}.  These deposits are older than the 1443 $\pm$ 7 Ma Purcell lavas and further constrained in age by a tuff with a U-Pb date of 1454 $\pm$ 9 Ma within the Helena Formation \citep{Evans2000c}. 
\item The Mesoproterozoic Iqqittuq Formation of the Borden basin (formerly part of the Society Cliffs Formation) contains bedded gypsum deposits (massive and laminated with beds that reach a thickness of 2.5 meters) and shale with halite casts \citep{Kah2001a}.  These deposits are bracketed between Re-Os dates of 1048 $\pm$ 12 Ma for an underlying shale and 1046 $\pm$ 16 Ma for an overlying shale \citep{Gibson2018a}.
\item The Tonian Ten Stone Formation of the Mackenzie Mountains Supergroup (formerly known as the Gypsum Formation) contains a $\sim$500 meter thick succession dominated by gypsum with minor anhydrite interpreted to have been deposited in a deep-water (below wave base) restricted marine basin \citep{Turner2016a} . These thick bedded sulfate deposits are older than cross-cutting 777.7 $\pm$ 2.5 Ma sills of the Gunbarrel large igneous province (U-Pb date from \citealp{Jefferson1989a}) and younger than ca. 1005 Ma detrital zircons \citep{Turner2016a}. The overlying Ram Head Formation has been correlated with the Bitter Springs Stage which is constrained between 811.5 $\pm$ 0.3 Ma and 788.7 $\pm$ 0.2 Ma \citep{Macdonald2010a, Swanson-Hysell2015a} suggesting that the evaporites are ca. 820 Ma \citep{Turner2016a}. These deposits are hypothesized to be correlative with sulfate evaporites within the Minto Inlet Formation of the Shaler Supergroup \citep{Jones2010a, Turner2016a}.
\item The Tonian Kilian Formation of the Shaler Supergroup contains nodules of gypsum and anhydrite interpreted to have been deposited in an intertidal to supratidal evaporitic mudflat environment \citep{Prince2014a}. The Kilian Formation is interpreted to post-date the Bitter Springs Stage and be correlative with the Redstone River Formation of the Coates Lake Group in the McKenzie Mountains that contains bedded gypsum as well as gypsum-bearing siltstone\citep{Jefferson1989a, Jones2010a}. The Redstone River Formation is younger than the 777.7 $\pm$ 2.5 Ma volcanics and older than a 732.2 $\pm$ 4.7 Ma Re-Os isochron from the overlying Coppercap Formation \citep{Rooney2014a}.
\end{itemize}

@@

\subsection{Evaluating Laurentia's Proterozoic paleogeographic neighbors}

Many different paleogeographic connections between Laurentia and other Proterozoic cratons that have been proposed and utilized in paleogeographic models both prior to and following the amalgamation of Laurentia's constituent Archean provinces to form Laurentia. This section does not seek to be comprehensive in this regard, but rather I seek to highlight and contextualize some of the more prominent and/or well-supported models -- particularly those following the assembly of Laurentia.

\textit{Note that this section is currently under development and is incomplete}

\subsubsection{Associations prior to Laurentia assembly}

Within this volume, \cite{Salminen2021a} describes proposed Neoarchean to Paleoproterozoic supercratons (terminology of \cite{Bleeker2003a}) prior to Laurentia assembly. In particular, they discuss the hypothesis of Superia wherein the Superior province of Laurentia is central to a group of Archean lithospheric blocks \citep{Bleeker2006a}, including the Kola and Karelia provinces of Baltica and the Wyoming province of Laurentia, that broke up prior to 2.0 Ga. This hypothesis has largely been based on proposed shared connections in the source of mafic intrusive rocks from 2.45 to 2.11 Ga with Kola and Karelia on the southeast margin of the Superior Province in present-day coordinates \citep{Davey2020a}. While \cite{Salminen2021a}  argue that the originally proposed Superia of fit is not consistent with Baltica paleomagnetic poles, they propose a Superia (II) fit where there is a connection between the blocks between 2.68 and 2.05 Ga but with some internal rotations. Another proposed connection, also evaluated by \cite{Salminen2021a}, based on the correlation of mafic dikes is one between the Slave province of Laurentia and the Dharwar province of India \citep{French2010a} as part of Sclavia \citep{Bleeker2003a}. This Slave-Dharwar connection is found to be consistent with paleomagnetic data if modified into a Sclavia (II) orientation and these blocks would have had a distinct dirft history from Superia for most of the Paleoproterozoic \citep{Salminen2021a}.

@@

In the central and southern Appalachians there are inliers of rocks that were metamorphosed during the Ottawan phase of the Grenvillian orogeny \citep{McLelland2013a}. On the basis of whole-rock Pb-isotope data, \cite{Loewy2003a} and \cite{Fisher2010a} proposed that these inliers are fragments of Amazonia lithosphere that were transferred to Laurentia during the orogeny and left behind when the Iapetus Ocean formed. In particular, \cite{Fisher2010a} suggest the Suns\'as orogen of Amazonia as the best match for southern and central Appalachian inliers. This positioning leads to a paleogeographic model wherein Amazonia is a major portion of the conjugate continental lithosphere that collided with Laurentia during Rodinia assembly \citep{Evans2013b, Cawood2017a}. While the lack of ca. 1100 and ca. 1000 Ma poles from Amazonia precludes a robust paleomagnetic test, this scenario is consistent with the available late Mesoproterozoic poles from Amazonia (ca. 1200 Nova Floresta pole and ca. 1150 Fortuna Formation pole; D'Agrella-Filho et al., this volume) as shown in \cite{Evans2013b}. In this paleogeographic scenario, the basement inliers of the Appalachian Orogen in the Blue Ridge region are interpreted to be the leading edge of Amazonia with initial collision ca. 1080 Ma initiating the Ottawan phase of the Grenvillian orogeny (Fig. \ref{fig:tectonic_history}). Subsequent separation of Amazonia would have lead to the formation of the Iapetus Ocean as Rodinia rifted apart. Departure of Amazonia potentially occurred as early as ca. 700 Ma in the Paleo-Iapetus Ocean model of  \citet{Robert2020a}.
